The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-07-06 manuscript.Rmd
many edits to manuscript

Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize. The digitization of natural history collections and other biodiversity information systems can now be used to enable synthesis and highlight knowledge gaps. Here we use these data to synthesize the state of our knowledge for marine invertebrates in the continental US. We show that current knowledge and sampling efforts are inadequate given the diversity of the fauna and the geographic extent of US. While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value would be greatly improved by anchoring the results into existing DNA sequence data and associated vouchers. However, the best documented phyla have at most 25% of their species sequenced for the most widely used DNA barcoding gene, cytochrome oxidase I (COI). More sampling, coordinated at the scale of the US, is needed to improve our knowledge of species and their geographic ranges. This data is needed to detect the impact of climate changes and other anthropogenic strains to global environments. Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.

@@

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010]. For marine organisms, because of the diversity of their complex life history strategies and their size range, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016], making it difficult to monitor changes in species distributions and abundances. While most documented global extinctions of marine species are vertebrates, many others may have been lost unnoticed because they were not known [@Sala2006]. There is therefore a pressing need to improve our knowledge of marine systems before their functioning is irreversibly altered, and species go extinct.

@@

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions of global research. We show that public databases are relatively incomplete, and that most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances. This can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.

Here, we focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010]. The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest on the West coast. With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas. While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast. This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity databases. However, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.

@@

Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area. We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1987]. Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r format(nrow(idigbio_gom_records), big.mark=",")` and `r format(nrow(obis_gom_records), big.mark=",")` records for iDigBio and OBIS respectively, representing `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species. For the Pacific Northwest, we retrieved `r format(nrow(idigbio_pnw_records), big.mark=",")` and `r format(nrow(obis_kozloff_records), big.mark=",")` records for iDigBio and OBIS respectively representing `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species. For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-overlap)). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)). Previous esimates for species richness in the Gulf of Mexico for marine invertebrates were of `r format(11150-1541-434, big.mark=",")` species, and `r format(8374-909-134, big.mark=",")` for the Pacific coast (including California) [@Fautin2010].

```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right). This plot is generated using the UpSet technique, which uses a matrix-based layout to represent intersections of set and their sizes. The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black filled circle in the matrix cell. Black lines connect the sets that are part of the intersections. (\\#fig:database-coverage)', echo=FALSE}
knitr::include_graphics("figures/database_coverage.pdf")
```

We investigated the number of species for each phylum that contributed to the discrepancies between the regional species lists and the database records. We found that the better studied phyla contributed the most to the differences across the data sources. OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig. \@ref(fig:database-overlap))

Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.

@@

Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are preserved less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of its range: it was included in the list because it is known to occur at the limit of its geographic zone, but has never been collected within the boundaries; (5) does not have geographic coordinates associated with it: it can be in the database but without georeferenced datapoints that made it possible to discover them. By querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that dok `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% fall within the US waters (n = `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`).

Species that are in the database but not in the list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erroneously excluded the records (2) records that have been discovered since the list has been assembled (for iDigbio records from the Gulf of Mexico, this represent `r format_output(mean(recently_collected$gom_in_idig$min_year > 2001))`%, n=`r nrow(recently_collected$gom_in_idig)` and `r format_output(mean(recently_collected$pnw_in_idig$min_year > 1987))`%, n=`r nrow(recently_collected$pnw_in_idig)`); (3) records with incorrect identification; (4) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.

@@

To assess whether species included in regional lists but absent from the same geographic area nevertheless occur elsewhere, we queried the iDigBio and OBIS databases for all the marine invertebrate species included in these lists.

We extracted `r format(sum(gom_worms$is_marine, na.rm = TRUE), big.mark=",")` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` species from the Pacific Northwest. We then looked for the number of species listed that are known to occur in the same geographic area, the United States, or anywhere else in the world. Overall, OBIS had georeferenced records for most of the species that are included in the species lists. However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.

>>>>

Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize. The digitization of natural history collections and other biodiversity information systems can now be used to enable synthesis and highlight knowledge gaps. Here we use these data to synthesize the state of our knowledge for marine invertebrates in the continental United States. We show that current knowledge and sampling efforts are inadequate given the diversity of the fauna and the geographic extent of US. While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value will be greatly improved by anchoring the results into existing DNA sequence data and associated vouchers. However, the best documented phyla have at most 25% of their species sequenced for the most widely used DNA barcoding gene, cytochrome oxidase I (COI). More sampling, coordinated at the scale of the US, is needed to improve our knowledge of species and their geographic ranges. This data is needed to detect the impact of climate changes and other anthropogenic strains to the environment. Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.

@@

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010]. For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016], making it difficult to monitor changes in species distributions and abundances. While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006]. There is therefore a pressing need to improve our knowledge of marine systems before their functioning are irreversibly altered, and species go extinct.

@@

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions in global research. We show that while public databases are capturing a significant portion of our knowledge about species richenss, most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but more comprehensive data on species distributions and abundances are needed. These goals can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.

Here, we focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010]. The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic shores of the Pacific Northwest on the West coast. With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas. While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast. This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity databases. Additionally, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.

@@

Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area. We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1974], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` species respectively. Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. We filtered the results to only include the same higher taxonomic level taxa as in the faunistic lists, and kept only records for which the species could be matched to a species name in WoRMS. For the Gulf of Mexico, we retrieved  `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species in iDigBio (from $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records), and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_gom_records), big.mark=",")` records). For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")`) and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_kozloff_records), big.mark=",")`). Overall, the number of species known from the Gulf of Mexico match the number estimated by experts (`r format(11150-1541-434, big.mark=",")` species [@Fautin2010]). For the Pacific coast, this number is lower than the estimate (`r format(8374-909-134, big.mark=",")` species [@Fautin2010]), but this estimate included species from California and from deeper areas that were not included in Kozloff's faunistic list.

For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-coverage)). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)). Actually, the majority of the species in the faunistic list are not recorded in any of the public databases. Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.

```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right). This plot is generated using the UpSet technique[@Lex2014], which uses a matrix-based layout to represent intersections of sets and their sizes. The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black-filled circle in the cells underneath. Black lines connect the sets that are part of the intersections. (\\#fig:database-coverage)', echo=FALSE}
knitr::include_graphics("figures/database_coverage.pdf")
```

We investigated the number of species for each phylum that contributed to the discrepancies between the regional species lists and the database records. We found that the better studied phyla contributed the most to the differences across the data sources. OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig. \@ref(fig:database-overlap)). Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have not digital records for them in these geographic areas.

@@

Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced datapoints and therefore were not returned in our queries.

Untangling the relative contribution of these factors to explain the discrepencies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepency, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico.

Species that are in the database but not in the faunistic list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erroneously excluded by the author of the checklist; (3) records that have been discovered since the list has been assembled; (4) records with incorrect identification: records identified as a species that does not occur in the area; (5) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.

We investigated the proportion of species discovered since the publication of the faunistic lists using iDigBio records. They represent `r sum(recently_collected$gom_in_idig$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_idig$min_year > 1987)` species for the Pacific Northwest. While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the presence of these species based only on museum records, as these lists were mostly assembled from publications, and only a small fraction of museum records are used in publications.

@@

To assess whether species included in regional faunistic lists but absent from the geographic area in public databases nevertheless occur elsewhere, we queried the iDigBio and OBIS using the species names included in these faunistic lists.

We extracted the `r format(sum(gom_worms$is_marine, na.rm = TRUE), big.mark=",")` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` species from the Pacific Northwest. We then looked for the number of species from these faunistic lists that are known to occur within the geographic area, or somewhere else in the world. Overall,
OBIS had georeferenced records for most of the species that were included in the faunistic lists for the Gulf of Mexico. However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.
